
This thesis is the assignment for the course MA3911 Master Thesis in Mathematical Sciences.

\section*{Acknowledgments}
I want to thank Professor Robert Brian O'Hara and Dr. Ron Ronald Togunov for supervising this thesis. I am particularly grateful to Dr. Ronald Togunov for consistent meetings and feedback throughout this project. I also want to thank Simen Furseth Knutsen for suggestions on how to reduce bias in the estimates of diffusion processes. Lastly I want to thank my sister for spending time reviewing my thesis.


\section*{Relevancy to UN Development Goals}
Studying animal habitats is an important part of species conservation. As we move into a future where the environment is increasingly affected by human pollution, studying animal behavior becomes more important to combat loss in biodiversity. The UN sustainability goals include goals to "combat climate change and its impacts", "Conserve and sustainably use the oceans, seas and marine resources for sustainable development" and "Protect, restore and promote sustainable use of terrestrial ecosystems, sustainably manage forests, combat desertification, and halt and reverse land degradation and halt biodiversity loss"\parencite{UN_goals}. This thesis is relevant to these goals, since it explores how to improve utilization distribution estimates. Improved utilization distribution estimates can be used to make informed decisions to ensure the future of endangered ecosystems.







\newpage
\null
\thispagestyle{empty}
\newpage